
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Expectations}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{itemize}
\item Please note, the use of cell phones is not permitted during lecture.  
These should be left in your backpack or not brought to class. 
\item Laptop computers may be used for taking notes or for in-class assignments but cannot be used for other activities during lecture.
\item No student may record any classroom or laboratory activity without the express written consent of the instructor. 
If a student believes that he/she is disabled and needs to record or tape classroom activities,
 he/she should contact the Office of Equal Opportunity Services, Disability Resources to request an appropriate accommodation. 
 \end{itemize}
 
Please keep in mind that these guidelines are necessary to maintain an environment that is safe and conducive for learning.
 
\subsection{Expectations of Students and Instructors}
The instructors and teaching assistants have the right to expect the following of students:
(Adapted and modified from those developed by \href{http://home.snu.edu/~HCULBERT/contract.htm}{Howard Culbertson at Southern Nazarene University}).
\begin{enumerate}
\item Students will arrive to class on time and will be prepared for the lecture.
\item Students will turn in assignments on time (see policy on assignment due dates).
\item Students will immediately inform the instructor or the teaching assistants if extenuating circumstances prevent the student from attending a lecture.
\item Students will follow the code of conduct regarding academic integrity, cheating, plagiarism, and collaboration as outlined in the syllabus.
\item Students will seek assistance when they need it.
\item If contacted by the instructor or teaching assistant, students will respond within 24 hours during the week and 48 hours on weekends.
\end{enumerate}
The students have the right to expect the following of the instructor and teaching assistants:
\begin{enumerate}
\item A syllabus that describes class procedures, policies, and a course description will be provided.
\item Class sessions that will start and end on time.
\item Any changes to the course schedule will be provided to the students within 48 hours of the change.
\item The instructor will be available outside class either during their posted office hours or during other pre-arranged times.
\item If contacted by a student, the primary instructors or teaching assistants will respond within 24 hours during weekdays and 48 hours on weekends.
\end{enumerate}
