%% adapted from Chris Bourke's syllabus template for CS 1
%% https://github.com/cbourke/ComputerScienceI
%% Accessed on 29 July 2020
%% Used and distributed under the CC BY-SA 4.0 License

\documentclass[12pt]{scrartcl}
%\usepackage{tagpdf}

\usepackage{epsfig,amssymb} 

%\usepackage{draftwatermark}
%\SetWatermarkScale{7}

\usepackage{xcolor}
\usepackage{graphicx}
\usepackage{epstopdf}
\usepackage{multirow}
\usepackage{colortbl} 
\usepackage{xspace}
\usepackage[normalem]{ulem}
\usepackage{multicol}
\setlength{\marginparwidth}{2cm}
\usepackage{todonotes}

\usepackage{tcolorbox}

\definecolor{steelblue}{RGB}{70, 130, 180}
\definecolor{darkred}{rgb}{0.5,0,0}
\definecolor{darkgreen}{rgb}{0,0.5,0}
\usepackage[pdflang={en-US}]{hyperref}
\hypersetup{
  colorlinks,
  linkcolor=darkgreen,
  citecolor=darkgreen,
  menucolor=darkred,
  urlcolor=blue,
  pdfpagemode=UseNone,
  pdftitle={Syllabus},
  pdfauthor={Dan DeBlasio},
  pdfkeywords={}
}
\setcounter{tocdepth}{2}

\usepackage{fullpage}
\pagestyle{empty} %
\usepackage{subfigure}
\usepackage{enumitem}
\setenumerate{nolistsep}
\setitemize{nolistsep}
\renewcommand{\labelenumii}{\alph{enumii}.}


\setlength{\parindent}{0pt} %
\setlength{\parskip}{.25cm}
\usepackage{lastpage}
\usepackage{fancyhdr}
\renewcommand*{\titlepagestyle}{fancy}
\pagestyle{fancy}
\setlength{\headheight}{14.5pt}
\addtolength{\topmargin}{-14.5pt}
\renewcommand{\headrulewidth}{0.0pt}
\renewcommand{\footrulewidth}{0.4pt}

\lhead{~}
\chead{~}
\rhead{~}
\lfoot{\Title ---Syllabus}
\cfoot{~}
\rfoot{\thepage\ / \pageref*{LastPage}}

\makeatletter
\title{\large Essential Mathematics and Statistics for Scientists}\let\Title\@title
\subtitle{
{\small
Carnegie Mellon University\\
School of Computer Science\\
Computational Biology Department
}
\vskip-1cm}
\date{\large 02-680 -- Fall 2023\\ \vspace{1em}\small Revised: \today\\(change since start of classes marked \change{}{in orange})}
\author{}
\makeatother

\newcommand{\change}[2]{\textcolor{orange}{#2}}
%\newcommand{\change}[2]{#2}


%\tagpdfifpdftexT
% {
%  \pdfcatalog{/Lang (en-US)}
%  \usepackage[T1]{fontenc}
% }
% 
%\tagpdfsetup{activate-all,tabsorder=structure}

\begin{document}
%\tagstructbegin{tag=Document}

\maketitle

%\begin{center}
%{\Huge\color{red}DRAFT}
%\end{center}



%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\section{General Information}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\tagmcbegin{tag=P}
%\paragraph{Course Number:} 02-680

\paragraph{Class Location and Times:} Posner Hall (POS) 153,\\
\hspace*{3em}Tuesday \& Thursday 3:30pm--4:50pm (Lecture), \\
\hspace*{3em}Friday 5:00pm-6:20pm (Recitation)



\paragraph{Course Description:} 
This course rigorously introduces fundamental topics in mathematics and statistics 
to first-year master's students as preparation for more advanced computational coursework. 
Topics are sampled from information theory, graph theory, proof techniques, phylogenetics, 
combinatorics, set theory, linear algebra, neural networks, probability distributions and densities, 
multivariate probability distributions, maximum likelihood estimation, statistical inference, hypothesis testing, 
Bayesian inference, and stochastic processes. 
Students completing this course will obtain a broad skillset of mathematical techniques and 
statistical inference as well as a deep understanding of mathematical proof. 
They will have the quantitative foundation to immediately step into 
an introductory master's level machine learning or automation course. 
This background will also serve students well in advanced courses that apply concepts in 
machine learning to scientific datasets, such as 
02-710 (Computational Genomics) or 02-750 (Automation of Biological Research). 
The course grade will be computed as the result of homework assignments, midterm tests, and class participation.

%\tagmcend

\paragraph{Prerequisite:} 
High school mathematics (including single- and multi-variable calculus) and ability to reason mathematically.

\paragraph{Course Goals:} 
Students completing this course will have a broad overview of mathematical techniques, 
as well as a deep understanding of proof techniques and statistical tests. 
They will have the quantitative foundation to immediately step into a machine learning or 
automation course aimed at computational biology or other science MS students. 
This background will also serve them well in advanced courses that apply concepts in machine learning 
to scientific datasets, such as 02-710 (Computational Genomics).

\clearpage
\tableofcontents



\input{instructional_staff}

\clearpage
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Logistics}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\paragraph{Course Topic Schedule:} Can be found \href{https://docs.google.com/spreadsheets/d/1VhhVtQ6OiOiUol-YNR7yw1RdQCwQw4jsiwoejJQ6trk}{on Google Sheets}\footnote{\url{https://docs.google.com/spreadsheets/d/1VhhVtQ6OiOiUol-YNR7yw1RdQCwQw4jsiwoejJQ6trk}}
and is subject to change.


\paragraph{Topics covered this semester:}

\begin{multicols}{2}
\begin{itemize}
%\item Information theory** % Not currently in course schedule
%\item Graph theory** % Not currently in course schedule
%\item Proof techniques* % Not explicitly in the outline but hard to avoid (probably not this high on the list)
%\item Phylogenetics* % Not explicitly in the outline
%\item Combinatorics** % Not currently in course schedule
%\item Set theory* % Not explicitly in the outline but implicity
\item Sets and functions
\item Linear algebra
\item Matrices and vectors
%\item Neural networks** % Not currently in the course schedule
\item Probabilities
\item Probability distributions 
\item Probability densities
\item Multivariate distributions
\item Maximum likelihood estimation
\item Statistical inference
\item Hypothesis testing
\end{itemize}
\end{multicols}

\paragraph{Textbook:} 
There are no required textbooks for this course. Below are textbooks that can be good references: 
\begin{itemize}
\item \textbf{Mathematics for Machine Learning} by Deisenroth, Faisal, Ong
\item \textbf{All of Statistics: A Concise Course in Statistical Inference} by Larry Wasserman 
\item \textbf{Probability and Statistics} by Morris DeGroot and Mark Schervish
\item \textbf{Linear Algebra Done Right} by Sheldon Axler
\end{itemize}

\paragraph{Communication platforms:}
\begin{itemize}
\item \textbf{Canvas Homepage.} 
The course homepage will be hosted on Canvas. 
Canvas will be used for attendance and as a central repository for grades. 
You should be automatically enrolled at \url{https://canvas.cmu.edu/courses/36680}.
\item \textbf{Discussion Forum.} 
An online forum is provided on Piazza as an area for discussion and questions. 
The forum will be moderated by the course staff who will respond to questions, 
but students are encouraged to help each other via discussion. 
However, assignment specifics should not be discussed --- 
any hints will be provided by the teaching staff. 
You can find the class on Piazza at\\ \url{https://piazza.com/cmu/fall2023/02680}.
\end{itemize}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Grading}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Grades are communicated to students in a timely manner. 
It is the students’ responsibility to keep track of their grades by compiling the grades they receive. 
Your semester grade will be based on a combination of homework assignments, weekly quizzes, class participation,  mid-term assessment, and a final exam. 

The approximate percentages are as follows:
\begin{center}
\begin{tabular}{rl}
\textbf{50\% } & Homework\\
\textbf{10\% } & Attendance \& Participation\\
\textbf{40\% } & Exams (20\% per)\\
\end{tabular}
\end{center}

The base percentage-score-to-letter-grade conversion for this course is as follows: 

\begin{center}
\begin{tabular}{r|ccc}
 & + & & -\\
 \hline
A & 98\% & 93\% & 90\%\\
B & 88\% & 83\% & 80\%\\
C & 78\% & 73\% & 70\%\\
D & 68\% & 63\% & ---\\
\end{tabular}
\end{center}
Any lower scores will result in a grade of ``R''.
These minimums may be lowered without notice but will not be raised. 

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Homework}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Written homework assignments will test your knowledge of the material covered in class. 
Homework assignment will be made available at least one week before the due date.

All writing assignments are to be submitted on Canvas on the day it is due.  
Any assignment turned in after the due date will be deducted 5 percentage points for the first day it is late, 
and 20 percentage points for every day after that.
If you have extenuating circumstances, just communicate with us and we'll do what we can to accommodate.  

Homework will be submitted as a \texttt{pdf} document which will need to be generated using an editor of the students choosing. 
Due to the reliance of this course on mathematical notation the use of \LaTeX{} is encouraged, 
If using another editor, students should utilize the editor's equation editor to ensure the correct answers are being communicated. 
To assist those students using \LaTeX, the source code for each assignment will be provided in addition to the \texttt{pdf}.

\begin{tcolorbox}[colback=red!5,colframe=red!75!black,title=Cheating policy]
All work must be your own. Unauthorized collaboration or \href{https://www.cmu.edu/student-affairs/ocsi/}{plagiarism} will result in a negative grade (e.g., a homework worth 100 points will be factored in as a -100 points towards your final grade) and will be reported to your academic advisor and dean.

\vspace{1em}
See Section~\ref{sec:cheating}.
\end{tcolorbox}

\subsubsection{Regrading of Assignments}
In the event that you feel an assignment was graded incorrectly, 
you must make the request in writing within one week of receiving the graded material.  
Your request should include a reference to the section of the assignment you believe was incorrectly graded.   
By asking for your assignment to be regraded, 
the whole assignment will be regraded and it is possible that fewer points 
may be awarded than the previous grade in light of new consideration.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Attendance \& Participation}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Attendance will be taken, 
and we will have occasional in-class exercises that serve to reinforce the concepts we have covered. 
These exercises will not be graded, but participation will be expected in order to receive a complete grade for that day.
You are allowed three ``dropped'' attendance grades without penalty. 
These can be used for any purpose. 

All classes will be live.
All efforts will be made to record lectures, but will only be made available in extreme circumstances.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Exams}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
The two exams will test your knowledge of the material from the class. 
The first midterm will be held in class, and the second midterm will be held during the university’s scheduled time. 
Dates for the exams will be announced in class, and can be found on the course schedule when available.

The exams will not be cumulative: midterm 2 will cover material encountered after midterm 1. 
That having been said, later material in the class may build upon the techniques covered in the earlier material.


\input{expectations}

\input{integrity}

\input{resources}

\end{document}